\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{enumerate}
\usepackage{color}
\usepackage{mdwlist}
\usepackage{hyperref}

\setlength{\parindent}{0em}
\setlength{\parskip}{0.5em}

\title{\vspace{-0.75in}ASTR 535 Lab notes}
\author{Jon Holtzman}
\date{Spring 2016}

\begin{document}
\maketitle

\section*{Time, coordinate systems, observability tools}
\subsection*{Time Systems}
Systems of time: see \textcolor{blue}{\href{tycho.usno.navy.mil/systime.html}
        {Naval observatory reference}}
        for a full listing of different types of time.
\subsubsection*{Solar Time}
Time tied to position of Sun; based on amount of time it
takes for the sun to return to the same position in the sky
(aka days).
Note the distinction
between \emph{mean} solar time (clock time)
and \emph{apparent}
solar time (sundial, the ``equation of time'' and the analemma).

Most used solar time is Universal time.
UT = local mean solar time at Greenwich = ``Zulu''.
Tied to location of Sun, but average to ``mean sun''.

Local time: accounts for longitude of observer.
For practicality, legal time is split into time zones.

In detail, official time is kept by atomic clocks
(International Atomic Time, or TAI), and coordinated UT
(UTC) is atomic time with leap seconds added to compensate
for changes in earth's roation, where these are added to
keep UTC within a second of solar time (UT1).
See \textcolor{blue}
%{\url{https://en.wikipedia.org/wiki/Universal_Time}}
{\href{https://en.wikipedia.org/wiki/Universal_Time}
{here}} for some details.

\subsubsection*{Sidereal time}
Times based on position of stars, i.e. Earth's sidereal
rotation period $\sim$ 23h 56m 4s. Local sidereal time is GMST
(Greenwich mean sidereal time) minus longitude. At the vernal
equinox (time in sky when Sun crosses the celestial equator
as its declination is increasing), sidereal time = UT.
Difference between UT and GMST is one rotation (day) over
the course of a year, so about 2 hours per month.

Sideral is relevant for position of stars: stars come back
to the same position every sidereal day. As we'll see
below, \textcolor{red}{a given star crosses the meridian when the
local sidereal time equals the right ascension of the
star.}

\subsubsection*{Calendars}

Standard calendar is Gregorian, with leap years, etc.

For astronomy, it is simpler to keep track of days rather
    then year/month/day. Most dates given by the
    \href{https://en.wikipedia.org/wiki/Julian_day}
    {\textcolor{blue}{Julian date}}
    (number of days since UT noon, Monday,
    January 1, 4713 BC). Variations include modified Julian
    data (JD - 2400000.5 fewer digits and starts at midnight),
    heliocentric Julian date (JD adjusted to the frame of
    reference of the Sun, so can differ by up to 8.3 minutes).
    Heliocentric JD is the amount of time it would take a
    pulse of light to arrive at the sun.

Note that repeating events are often described as an event
    \emph{ephemeris:} $t_i(event) = t_0 + i(period)$.

The term \emph{ephemeris} is also used to describe how the
    position of an object changes over time, e.g. planetary
    ephemerides.

\subsection*{Coordinate systems}
LPL website on \textcolor{blue}
{\href{http://spider.seds.org/spider/ScholarX/coords.html}
{astronomical coordinate systems}}
\subsubsection*{Celestial coordinate systems}
\textcolor{blue}
{\href{http://csep10.phys.utk.edu/astr161/lect/time/coordinates.html}{(diagram)}}
\begin{itemize}
    \item RA-DEC: tied to Earth rotation, longitude and latitude.
        Zero RA at vernal equinox
    \item ecliptic: tied to plane of Earth rotation around the Sun.
        Zero ecliptic longitude tied to vernal equinox.
    \item galactic: tied to plane of the Milky Way
\end{itemize}

At vernal equinox, RA = 12h crosses the meridian at midnight.

Note that for a celestial coordinate system tied to the
Earth's rotation,
coordinates of an object change over time because of the changing direction
of the Earth's axis: precession and nutation. Because of this, coordinates are
always specified for some reference equinox: J2000/FK5, B1950, etc.; if using
coordinates to point a telescope, you need to account for this (but generally,
telescope software does this on its own). Note distinction between equinox and
epoch, where the latter is relevant for objects that move (which everything does
at some level).

Transformations between systems straightforward from spherical
trigonometry.

Note the common usage of an Aitoff projection (equal areas)
of the sky in celestial
coordinates, with location of ecliptic and galactic plane.
Software tools (Python, projection=``aitoff'' in subplot,
IDL: aitoff and aitoff\_grid in Astronomy users library).

\subsubsection*{Local coordinate systems}
\begin{itemize}
    \item Equatorial: HA-dec. $HA=LST - \alpha.\ LST=GMST - longitude$.
        Note normal convention for HA is to get larger to the west, i.e.
        opposite of RA. Objects at zenith have $\delta=$ latitude of observer.
    \item Horizon: alt-az or zd-az
\end{itemize}
Local coordinates are important for pointing telescopes. Note that there are
various other effects that one has to consider for pointing a telescope at a
source of known celestial position: proper motion, precession, nutation,
``aberration of light'', parallax, atmosphereic refraction.

\subsection*{Finding positions of celestial objects}
\begin{itemize}
    \item \href{http://simbad.u-strasbg.fr/simbad/}
        {\textcolor{blue}{SIMBAD}}:
        look up coordinates of many objects outside solar system
        by name, etc., also provides much other reference information.
    \item \href{http://vizier.u-strasbg.fr/viz-bin/VizieR}
        {\textcolor{blue}{VizieR catalog database}}
        Database of astronomical catalogs, with search and download
        possibilities.
    \item \href{http://ned.ipac.caltech.edu}
        {\textcolor{blue}{NED}}:
        NASA extragalactic database: galaxies, etc.
    \item solar system ephemerides: JPL
        \href{http://ssd.jpl.nasa.gov/horizons.cgi}
        {\textcolor{blue}{HORIZONS}}
\end{itemize}

\subsection*{Orientations of objects in the sky}
Usually specified by position angle: angle of object in degrees from NS line,
measured counterclockwise.

An important observational position angle for spectroscopy:
\emph{parallactic angle}, the position angle of the line from zenith to
horizon.

\subsection*{Observability}
In general, one would like to observe objects through the shortest
possible path through Earth's atmosphere, i.e., when they are \emph{transiting}
(crossing the meridian, HA=0). The more atmosphere the light goes through,
the more losses due to atmospheric absorption/scattering (more severe at
shorter wavelengths), and the more image degradation from atmospheric seeing.
Of course, it doesn't make sense to wait for an object to transit if you
don't have anything else to do in the meantime; efficient use of telescope time
is the primary concern.
One \emph{airmass} is the amount of air directly above an observer.
If you are looking at the zenith, you are looking through one airmass.
Generally, most observers attempt to observe at airmasses
less than 2, i.e.\ within 60 degrees of zenith. Once you hit an airmass
of 3, the object is rapidly setting (except at very high declination).
Of course, for some solar system objects (objects near the sun), one has no
choice but to observe at high airmass.

Note that HA gives some indication of observability, but that
higher declination objects can be observed to higher HA than lower
declination objects. Roughly, at the celestial equator, an HA of 3 hours
is about an airmass of 2, and in many cases, one doesn't want to go much
lower in the sky.

Another issue with observability has to do with the Moon,
since it is harder to see fainter objects when the sky is brighter.
Moon brightness is related to its phase, and to a lesser extent, to
distance from your object. Of course, if the Moon is below the horizon,
it does not have an effect. So for planning observations of faint objects,
one also has to consider Moon phase and rise/set times. Note that the
sky brightness from the Moon is a function of wavelength, and at IR
wavelengths, it is not a very significant contributor to the total sky
brightness; so often, telescopes spend bright time working in the IR.

\subsection*{Tools}
Here are some useful software tools to do tasks related to coordinate systems
and observability, though there are others out there. Anything that
accomplishes the desired tasks adequately is fine to use; just make sure
you're not limited by the tools that you choose. These are available
on the Astronomy Linux cluster; you can probably install them on your
laptop, but they will probably not be there by default.
\begin{itemize}
    \item \href{http://physics.dartmouth.edu/}
        {\textcolor{blue}{skycalc/skycalendar}}:
        text based programs, installed on our Linux cluster (link is
        to source code if you wish to install on your laptop).
        skycalendar gives daily almanac, position of moon, etc.
        skycalc allows you to enter coordinates of an object and
        obtain observability information for any specified date. Other
        features included as well: coordinate transformation, position
        of planets.
    \item \href{http://physics.dartmouth.edu/}
        {\textcolor{blue}{JSkyCalc}}:
        (java-jar /home/local/java/JSkyCalc.jar): JAVA implementation
        of skycalc, also installed on the Astronomy cluster (and
        available for download).
    \item \href{http://tdc-www.harvard.edu/wcstools/}
        {\textcolor{blue}{WCSTOOLS}}: full set of useful coordinate
        system programs, e.g.\ coordinate system transformation
        (command skycoor). Largely useful for use with coordinate
        system information in image
        headers (more later). Installed on the astronomy cluster.
    \item Python: \href{http://docs.astropy.org/en/stable/coordinates/}
        {\textcolor{blue}{astropy.coordinates}},
        IDL:
        \href{http://idlastro.gsfc.nasa.gov/ftp/pro/astro/euler.pro}
        {\textcolor{blue}{euler}} in Astronomy users library.
\end{itemize}


\subsection*{Exercises}
\begin{enumerate}[1.]
    \item \textbf{Predict the RA crossing the meridian at midnight for the
    first of every month. Try the command skycalendar (on the cluster,
    unless you download it yourself for your laptop) - give yourself a
    wide terminal window first - to see how well you did}

    \item \textbf{What time is it now? What is the sidereal time? What
    coordinates would it be most optimal to observe right now?}

    \item \textbf{When are the dark (no moon above horizon) first half
    nights in first quarter ? }

    \item \textbf{APO schedules the 3.5m in half-night blocks (A and
    B), split at midnight (or 1am during daylight savings). What are
    the best half-nights in the next year (month and half, e.g., Oct
    A, March B, etc.) to request to observe:}
        \begin{itemize}
            \item \textbf{Virgo cluster of galaxies (note central
            galaxy is M87, look up the coordinates)}
            \item \textbf{Galactic center (galactic coordinates are
            .... ask if you don't know!). You can use command skycoor
            (or Python or IDL tools) to convert galactic to equatorial
            (skycoor with no arguments gives syntax). }
            \item \textbf{Jupiter (look up its position using JPL
            HORIZONS) }
        \end{itemize}

    \item \textbf{Run skycalc (choose observatory A for APO, ? gives
    list of command help, look at r, d, y, and h commands). For the
    galactic center, what is the maximum amount of time it can be
    observed at an airmass of less than 2.5? How about the Virgo
    cluster? Why are these different?}

    \item \textbf{Run jskycalc. Play with all of the buttons! What
    planets will be visible spring 2016, and at what times of night?
    Note that you can load files with a list of coordinates, and you
    can make airmass observability charts for them.}

    \item \textbf{Start to outline plan for an 3 half-night observing
    run during late March A halves, when we are taking our APO trip.
    Eventually, the plan should include a list of objects for each
    night with a tentative order of observation, taking into account
    how much time needs to be spent on each object. Our projects are
    still TBD, but will likely include observations with multiple
    instruments.}
        \begin{itemize}
            \item \textbf{Determine the approximate range of RAs that
            we will be able to observe. }
            \item \textbf{Given the NMSU 1st quarter proposals, which
            of them might we be able to make some observations for?}
            \item \textbf{If you have other ideas for projects, start
            to tabulate them. (Sten/Diane stars for APOGEE
            calibration/neutron capture calibration, Triplespec RR
            Lyrae RV curves Drew Be stars)}
            \item \textbf{Start to prepare a joint web page with the
            plan, including relevant information: coordinates of
            objects, finder images if necessary, links to tabulated
            spectra, instrument manuals, etc. etc. }
        \end{itemize}

    \item \textbf{Look up the catalog Globular Clusters in the Milky
    Way in VizieR and download it (make sure to get all of the rows).}
        \begin{itemize}
            \item \textbf{Plot the locations in an Aitoff projection
            of equatorial coordinates. Can you detect Galactic
            structure?}
            \item \textbf{What clusters would be possible to observe
            during our March run? }
            \item \textbf{Convert coordinates to galactic coordinates
            and plot in an Aitoff projection.}
        \end{itemize}

\end{enumerate}

\newpage
\textcolor{magenta}{Friday, February 5}

\section*{Image display and graphical file-based display tools}
\subsection*{Image Display}
\subsection*{Quick introduction to astronomical image file format}
\subsection*{Standalone display tools}
\subsection*{Basic display operation}
\subsection*{Exercises}


\textcolor{magenta}{Friday, February 26}

\section*{Astronomical image processing packages: Introduction and
IRAF basics}

\subsection*{IRAF/DS9 basic operation}
\begin{itemize*}
    \item One time only: run \texttt{mkiraf}, which creates \texttt{login.cl}.
        This file can be customized at a later time if you have settings you want
        to start with every time.
        \begin{itemize*}
            \item to enable a larger frame buffer for display, uncomment and
                modify line: \texttt{stdimage = imt2048}
        \end{itemize*}
    \item The preferred method of running IRAF in the modern era is using the
        PYTHON interface, \texttt{pyraf}:
        \begin{itemize*}
            \item You can use pyraf via a normal python interface:

                \texttt{from pyraf import iraf}

                You will then need to use standard PYTHON styntax, rather
                than the old IRAF \texttt{cl} syntax.
            \item \texttt{pyraf} actually runs a front-end interpreter to
                emulate the original IRAF command-line interface. This is
                convenient for previous users and for some tasks, but
                ``hides'' the Python interpreter and its power.
        \end{itemize*}
    \item For displaying images, start an image display tool, i.e.\ DS9,
        in the background: \texttt{ds9 &}. Be aware that the stdimage that is set
        in the login.cl file may limit the maximum size of the image that will
        be displayed.
    \item help:
        \begin{itemize*}
            \item there is an internal \texttt{.help} command.
            \item IRAF help
            \item tutorials
        \end{itemize*}
    \item basics:
        \begin{itemize*}
            \item IRAF contains many programs for astronomical analysis. These
                are grouped into \emph{packages}, and individual commands are
        \end{itemize*}
\end{itemize*}


























\end{document}
