\documentclass[12pt]{article}
\usepackage[margin=0.75in]{geometry}
\setlength{\parindent}{0em}
\setlength{\parskip}{0.75em}
\usepackage{fancyhdr}
\pagestyle{fancy}
\setlength{\headheight}{15pt}
\lhead{Laurel Farris}
\chead{ASTR 535 --- Exposure Time Calculator paragraph}
\rhead{06 May 2016}

%------------------------------------------------------------------------------%
\begin{document}

The first thing I did for this project was send a weekly report in pdf form.
My initial task for the exposure time calculator was to write the module
for calculating atmospheric extinction. I had everything written out, but still
needed a file for wavelength-dependent extinction coefficients. I think
another student said they would look for that, but I wasn't sure.
I set up a github account where all the codes, reports, documentation, etc.\
could be kept in one place for everyone to access, if they didn't know how
to fork or otherwise interact in the ``git'' manner.

Overall I think this was a good assignment, and I definitely appreciate the
importance of being able to work in a group liek that. I don't think the group
collaboration itself
went very well at all though. I thought the e-mail chains were a terrible way to
communicate (I always miss things when they're buried in the replies, so I
was late to one meeting and completely missed another)
but no one else seemed interested in finding an alternative.
I volunteered to do a few things, but then someone else would do them, so I
got a very confused as to what exactly everyone was expecting me to do.
We should have been more clear and specific when naming our modules, but
they were all called some variation of ``exp\_time\_calc''
or ``etc'' or ``ExposureTimeCalculator'', so I couldn't tell by looking at the
code directory what each module actually did. Documentation was also scarce;
I made a directory for this on the github account (which did contain a
README file explaining what the directories were for, but I'm not sure if
anyone actually read that).
Another student made a different github account for the codes because apparently
mine was ``too cluttered'' but no one said a word to me about re-organizing it
or getting rid of uneccessary files. Now we have two directories, which
defeated my purpose of keeping everything in one place.
I put a lot of work into organizing that directory for everyone, but
now I'm afraid that it looks like I didn't do much since a big part of
my contribution was kicked out. I'm not blaming any one particular student,
but the communication was really bad, including on my end, and I think
this is just due to lack of experience in group work (something my
introverted self has always been particularly bad at).

I think we really needed some more guidance, not on how to do the tasks themselves,
but on how to work together on a coding project.
I'm pretty sure not a single student in our class has
done that before. We could have benefited from some suggestions or useful
resources (github or otherwise) for communicating, storing and sharing codes,
dividing up tasks, setting deadlines, coming up with a general plan.
I really believe everyone tried their best, and I know it's important to figure
things out on our own on some level, but we have teachers for a reason, and
I think expecting us to figure out how to collaborate on top of the
assignment itself was a bit too much, and we really needed some help on that
aspect of the project.

Sorry this was a lot more than one paragraph.

%---------------------------------------------------------------------------%
\end{document}
